Approximately Exact Calculations for Linear Mixed Models

### 2015-05-22 ms.tex
implication of setting maxit = infty

@@

With \textbf{D1} and \textbf{D2} the \textcompute{R} function \textcompute{findf}, sketched below, will evaluate, within a box $B^0$, $\logRL$ to arbitrary accuracy everywhere $\logRL$ is large.  These notes explain some parts of the function.
\begin{enumerate}
\item A dataframe \textcompute{lines} is an input to \textcompute{findf}.  \textcompute{lines} contains the slopes
  and the intercepts on both axes of the $s_z+2$ lines discussed in Section~\ref{subsec:lines}.
  Details for computing them from $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ are in the Appendix.
\item \textcompute{startbox}, or $B^0$, is an input to \textcompute{findf}.  As explained at the beginning of
  Section~\ref{sec:algorithm}, \textcompute{startbox} could be $B_1$, $B$, or any other box the user chooses.
\item Constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are inputs to \textcompute{findf}.
	\begin{enumerate}[label=(\alph*)]
	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
		\textcompute{findf}'s loop.
	\item We will not further analyze regions of the plane where
		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B^0$
	         unless evaluation is stopped by one of the other criteria.
	\item We will not distinguish values of $\sigesq$ separated by less
		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
		less than $\delta_s$ (could be 0).  Separation may be specified
		in either absolute or relative terms. (I.e.\ we look at either the difference
		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
		side of the box to the other.)
	\end{enumerate}
\item A \textcompute{box} is a list consisting of
	 \begin{itemize}
	 \item upper and lower limits on $\sigesq$,
	 \item upper and lower limits on $\sigssq$,
	 \item upper and lower bounds on $\logRL$, $U^b$ and $L^b$ from \eqref{eq:bounds},
	 \item indicators for whether this box lies above (or to the right of), below (or to the left of),
	   or straddles each of the $s_z+2$ lines.
	 \end{itemize}
\item \textcompute{killfunc} is a function, shown below \textcompute{findf}, that determines whether
	a box $b$ should be divided more finely.
\item \textcompute{splitbox} is a function that takes a box $b =
	[\sigesq{}_\text{low}, \sigesq{}_\text{high}] \times [\sigssq{}_\text{low}, \sigssq{}_\text{high}]$ as input 
	and returns the four boxes created by dividing each side of $b$ at its midpoint.
%\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
%	and \eqref{eq:rll} for definitions and the Appendix for details --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
%	\dots, s_z+2$.
%\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
%\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
%	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
%\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
%	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
%	lower bound on $\logRL(\sshat)$.
%\item While the active list is not empty or we have not reached \maxit\
%	iterations
%	\begin{enumerate}
%	\item For each active box $b$:
%		\begin{enumerate}
%		\item if $U^b < L-M$ move $b$ to the inactive list.
%		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
%		\item if the horizontal extent of $b$ is less than $\delta_e$,
%			move $b$ to the inactive list
%		\item if the vertical extent of $b$ is less than $\delta_s$,
%			move $b$ to the inactive list
%		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
%			Remove $b$ and add the subboxes to the active list.
%		\end{enumerate}
%	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
%		lower bound on $\logRL(\sshat)$.
%	\end{enumerate}
%\item Return the active and inactive lists.
\end{enumerate}
\begin{verbatim}
findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
                  ratio = FALSE, lognote = "summary") {

>>>>

\subsection*{Alt: Lines Determine a Bounding Box}
%\label{sec:boundingbox}

Let $\sigesq{}^M$ and $\sigssq{}^M$ be the largest intercepts of the $s_z+2$ lines on the $\sigesq$ and $\sigssq$ axes, respectively.  I.e.,
\begin{align*}
  \sigesq{}^M &= \max \left\{ \frac{d_1}{c_1}, \cdots, \frac{d_{s_z+1}}{c_{s_z+1}} \right\}\\
  \sigssq{}^M &= \begin{cases}
                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}} \right\} &
                                  \text{if $\logRL$ is \eqref{eq:reexpress}}\\
                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}}, \frac{d_{s_z+2}}{c_{s_z+2}} \right\} &
                                  \text{if $\logRL$ is \eqref{eq:logpost}}.
                           \end{cases}
\end{align*}
Let $B_1$ be the box whose lower-left and upper-right corners are $(0,0)$ and $(\sigesq{}^M,\sigssq{}^M)$, respectively.  Figure~(\ref{fig:bigboundingbox}) illustrates with five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.\\
\textbf{Claim} \emph{All local maxima of $\logRL$ lie within $B_1$.}\\
\textbf{Proof} Let $p_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point such that $\sigssq{}_1 > \sigssq{}^M$ and let $\widetilde{p_1} = (\sigesq{}_1, \sigssq{}^M)$, as illustrated in Figure~(\ref{fig:bigboundingbox}).  For $j=1, \dots, s_z, s_z+2$, the partial derivatives \eqref{eq:drlde} are negative; for $j=s_z+1$, the partial derivative is zero.  Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$ and there can be no local maxima of $\logRL$ above the line $\sigssq = \sigssq{}^M$.  Let $p_2 = (\sigesq{}_2, \sigssq{}_2)$ be a point such that $\sigesq{}_2 > \sigesq{}^M$ and let $\widetilde{p_2} = (\sigesq{}^M, \sigssq{}_2)$.  For $j=1, \dots, s_z, s_z+1$, the partial derivatives \eqref{eq:drlds} are negative; for $j=s_z+2$, the partial derivative is zero.   Therefore, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and there can be no local maxima of $\logRL$ to the right of the line $\sigesq = \sigesq{}^M$.  \textbf{QED}\\

By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.  However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.  We will next find a box $B \supset B_1$ satisfying desideratum \textbf{D1}.  First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ by the same reasoning used to prove the previous claim.  But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.

%\begin{figure}[h]
%  \begin{subfigure}{.5\textwidth}
%	\centering
%	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
%	\caption{A rectangular region $B_1$ containing all maxima.}
%	\label{fig:bigboundingbox}
%  \end{subfigure}
%  \begin{subfigure}{.5\textwidth}
%	\centering
%	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
%	\caption{A smaller region containing all maxima.}
%	\label{fig:smallboundingbox}
%  \end{subfigure}
%  \caption{all local and global maxima lie in the regions bounded by the solid
%                dark line.
%               }
%  \label{fig:boundingbox}
%\end{figure}

\noindent\textbf{Claim} \emph{For any positive number $M$, there exist positive numbers $\widetilde{\sigma_e^2} > \sigesq{}^M$ and $\widetilde{\sigma_s^2} > \sigssq{}^M$ that determine a box $B \supset B_1$ whose lower-left and upper-right corners are $(0,0)$ and $(\widetilde{\sigma_e^2}, \widetilde{\sigma_s^2})$, respectively, such that
\begin{equation*}
  \sup_{B_1^c} \logRLss \le \logRL(\sshat) - M.
\end{equation*}}
I.e., \textbf{D1} is satisfied.\\
\noindent\textbf{Proof}
Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$.  Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.  Let $q_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point satisfying $\sigesq{}_1 \le \sigma_e^{2*}$ and $\sigssq{}_1 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line.  The partial derivatives (\ref{eq:drlds}), evaluated at $q_1$ are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but zero for $j=s_z+1$.  The partial derivatives (\ref{eq:drlde}), evaluated at $q_1$ are negative for $j \in \{1, \dots, s_z, s_z+1\}$ but nonnegative for $j=s_z+2$.  Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.  
\begin{equation}
	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
\end{equation}
which is larger than $M$ if
\begin{equation}
\label{eq:q1}
  \begin{split}
    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.

Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.  (Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy this criterion.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.  The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.  But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.

Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$ and let $q_3 = (\sigesq{}_3, \sigssq{}_3)$ satisfy $\sigesq{}_3 \ge \widetilde{\sigma_e^2}$ and $\sigssq{}_3 \ge \sigma_s^{2*}$.  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_3$ are nonpositive for $j \in \{1, \dots, s_{z+2}\}$; therefore, $\logRL(\sshat) - \logRL(q_3) \ge L - \logRL(\widetilde{q_3})$.  The latter expression is greater than $M$ iff $\logRL(\widetilde{q_3}) < L-M$.  But examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$ can be chosen large enough so that \textbf{D1} is satisfied.

Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$ and let $q_4 = (\sigesq{}_4, \sigssq{}_4)$ satisfy $\sigssq{}_4 \le \sigma_s^{2*}$ and $\sigesq{}_4 \ge \widetilde{q_4}$.  The partial derivatives \eqref{eq:drlds} evaluated at $q_4$ are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.  The partial derivatives \eqref{eq:drlde} evaluated at $q_4$ are nonpositive for $j \in \{1, \dots, s_z, s_z+2\}$  but nonnegative for $j=s_z+1$. Therefore,
\begin{equation*}
	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
\end{equation*}
which is larger than $M$ if
\begin{equation}
\label{eq:q4}
  \begin{split}
    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
  \end{split}
\end{equation}
Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.

Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.  \textbf{QED}\\
%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.  In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.  As already mentioned, if there is an $h$ line, then $q_1$ would have to be treated like $q_2$ and $q_3$.

%\begin{figure}[h]
%	\centering
%	\includegraphics[width=.5\linewidth]{figs/boundingbox2.pdf}
%	\caption{Box $B$ satisfies (\ref{eq:boundingbox})}
%	\label{fig:boundingbox2}
%\end{figure}

@@

With \textbf{D1} and \textbf{D2} the \textcompute{R} function \textcompute{findf}, sketched below, will evaluate, within a box $B^0$, $\logRL$ to arbitrary accuracy everywhere $\logRL$ is large.  These notes explain some parts of the function.
\begin{enumerate}
\item A dataframe \textcompute{lines} is an input to \textcompute{findf}.  \textcompute{lines} contains the slopes
  and the intercepts on both axes of the $s_z+2$ lines discussed in Section~\ref{subsec:lines}.
  Details for computing them from $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ are in the Appendix.
\item \textcompute{startbox}, or $B^0$, is an input to \textcompute{findf}.  As explained at the beginning of
  Section~\ref{sec:algorithm}, \textcompute{startbox} could be $B_1$, $B$, or any other box the user chooses.
\item Constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are inputs to \textcompute{findf}.
	\begin{enumerate}[label=(\alph*)]
	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
		\textcompute{findf}'s loop.
	\item We will not further analyze regions of the plane where
		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B^0$
	         unless evaluation is stopped by one of the other criteria.
	\item We will not distinguish values of $\sigesq$ separated by less
		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
		less than $\delta_s$ (could be 0).  Separation may be specified
		in either absolute or relative terms. (I.e.\ we look at either the difference
		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
		side of the box to the other.)
	\end{enumerate}
  Setting $\maxit = \infty$ and $\delta_e = \delta_s = 0$ implies that $B^0$ will be partitioned so that either
  $\logRL < \logRL(\sshat) - M$ or $\logRL$ is known to within $\epsilon$, for every $B^0_i$ in the partition.

\item A \textcompute{box} is a list consisting of
	 \begin{itemize}
	 \item upper and lower limits on $\sigesq$,
	 \item upper and lower limits on $\sigssq$,
	 \item upper and lower bounds on $\logRL$, $U^b$ and $L^b$ from \eqref{eq:bounds},
	 \item indicators for whether this box lies above (or to the right of), below (or to the left of),
	   or straddles each of the $s_z+2$ lines.
	 \end{itemize}
\item \textcompute{killfunc} is a function, shown below \textcompute{findf}, that determines whether
	a box $b$ should be divided more finely.
\item \textcompute{splitbox} is a function that takes a box $b =
	[\sigesq{}_\text{low}, \sigesq{}_\text{high}] \times [\sigssq{}_\text{low}, \sigssq{}_\text{high}]$ as input 
	and returns the four boxes created by dividing each side of $b$ at its midpoint.
%\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
%	and \eqref{eq:rll} for definitions and the Appendix for details --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
%	\dots, s_z+2$.
%\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
%\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
%	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
%\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
%	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
%	lower bound on $\logRL(\sshat)$.
%\item While the active list is not empty or we have not reached \maxit\
%	iterations
%	\begin{enumerate}
%	\item For each active box $b$:
%		\begin{enumerate}
%		\item if $U^b < L-M$ move $b$ to the inactive list.
%		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
%		\item if the horizontal extent of $b$ is less than $\delta_e$,
%			move $b$ to the inactive list
%		\item if the vertical extent of $b$ is less than $\delta_s$,
%			move $b$ to the inactive list
%		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
%			Remove $b$ and add the subboxes to the active list.
%		\end{enumerate}
%	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
%		lower bound on $\logRL(\sshat)$.
%	\end{enumerate}
%\item Return the active and inactive lists.
\end{enumerate}
\begin{verbatim}
findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
                  ratio = FALSE, lognote = "summary") {
